Numerical simulation and entropy dissipative cure of the carbuncle instability for the shallow water circular hydraulic jump

### 2020-11-19 numerics_paper/version2/numerical_chj.tex
Some notation changes, wording, and remove trailing whitespace.

\section{Numerical methods}
In this work, we consider Riemann-based finite volume methods.
Considering a one dimensional hyperbolic conservation law,
the cell average of the numerical solution given by a 
finite volume scheme (with forward Euler) can be written as follows:
\begin{align}\label{first-order_FV}
  \bfu_i^{n+1}=\bfu_i^n-\frac{\Delta t}{\Delta x}\left[\bfF_{i+1/2}-\bfF_{i-1/2}\right],
\end{align}
where $\bfF_{i+1/2}=\bfF_{i+1/2}(u^n)$ is a numerical flux; i.e., $\bfF_{i+1/2}$ is 
a discretization of the flux $\bff(\bfu(x_{i+1/2},t^n))$. 
This numerical flux is composed of two parts, a consistent term that approximates 
the physical flux at the interface between cells $i$ and $i+1$, and a stabilization term. 
For instance, the well-known Rusanov's numerical flux is given by 
\begin{align}\label{rusanov_num_flux}
  \bfF_{i+1/2}(\bfu)=\frac{\bff(\bfu_{i+1})+\bff(\bfu_i)}{2} - \lambda_{i+1/2}(\bfu_{i+1}-\bfu_i),
\end{align}
where $\lambda_{i+1/2}$ is an upper bound on the wave speed associated with the Riemann problem 
composed of the states $\bfu_i$ and $\bfu_{i+1}$. 
%If $z=u$ in \eqref{rusanov_num_flux}
%; i.e., if the Rusanov's numerical flux is obtained considering 
%the cell averages of the solution 
%and if the time step is small enough, 
Method \eqref{first-order_FV} with Rusanov's flux \eqref{rusanov_num_flux}
is positivity preserving and entropy stable, see for instance 
\cite{perthame1996positivity,tadmor2003entropy,guermond2016invariant}
However, the accuracy of the numerical flux, and hence of the method, is 
limited to first order. 
Two alternatives are common to improve the accuracy properties of Rusanov's method.
The numerical flux $\bfF_{i+1/2}$ can be approximated based on high-order polynomial 
reconstructions (or combinations of them) around the interface $i+1/2$. 
For smooth solutions, this strategy leads to arbitrarily high-order
methods. Alternatively, second-order accuracy can be obtained by removing some of the 
numerical dissipation introduced by the stabilization component of the numerical flux. 
The second-order Lax-Wendroff method \cite{lax1959systems} is a popular example of this approach. 
In this work we consider an adaptation of Lax-Wendroff method,
sometimes known as Lax-Wendroff-LeVeque \cite{leveque1997wave,leveque2002finite}, 
that incorporates limiters into the second-order correction. 
These high-order corrections to Rusanov's method require careful treatment to guarantee 
the solution is still positivity preserving and/or entropy stable.

The accuracy of a given first- or second-order method depends heavily on the Riemann solver. 
In particular, the Rusanov's Riemann solver is highly dissipative. In contrast, the well-known 
Roe's Riemann solver produces more accurate solutions. 
%
As we discussed in the introduction, finite volume methods with 
Roe's Riemann solvers are prone to the carbuncle instability for the CHJ. 
In contrast, if we use the Rusanov's 
Riemann solver, no carbuncles appear; however, many other important and physical features of 
the solution are dissipated, unless the grid is highly refined. 
Our main objective is to study the CHJ and to propose a Riemann solver
that is carbuncle free but do not dissipate important features of the solution. 
To do that we propose a Riemann solver that blends Rusanov's and Roe's solvers. 
Our blending criterion is based on the residual of the entropy (in)equality,
which induces dissipation of entropy near discontinuities. 
Blending these Riemann solvers has been proposed before
to solve the Euler equations, see e.g.
\cite{nishikawa2008very,wang2016developing,jaisankar2007diffusion,ohwada2018simple,deng2019new,ray2013entropy},
the shallow water equations, see e.g. \cite{bader2014carbuncle,kemm2014note},
and even the Navier-Stokes equations, see e.g. \cite{nishikawa2008very,ohwada2018simple}.
In these references, the authors use blending criteria different than ours.
Nevertheless, fixing the carbuncle instability by imposing entropy stability is not new.
For example, in \cite{ismail2009affordable,ismail2009proposed},
the authors developed carbuncle free solvers for the Euler equations
via dissipation of entropy.

@@

The second-order Lax-Wendroff-LeVeque method removes some of the numerical dissipation in 
\eqref{first-order_via_fluct} via a flux correction at each interface. The method is written as follows:
\begin{align}\label{second-order_via_fluct}
  u_i^{n+1} = u_i^n-\frac{\Delta t}{\Delta x}\left[\A^+\Delta Q_{i-1/2}+\A^-\Delta Q_{i+1/2}\right]
  -\frac{\Delta t}{\Delta x}\left[\tilde{F}_{i+1/2}-\tilde{F}_{i-1/2}\right],
\end{align}
where 
\begin{align*}
  \tilde{F}_{i\pm 1/2}=\frac{1}{2}\sum_p|s_{i\pm 1/2}^p|\left(1-\frac{\Delta t}{\Delta x}|s_{i\pm 1/2}^p|\right)\tilde\W_{i\pm 1/2}^p
\end{align*}
are the flux corrections at the interfaces $i\pm 1/2$. Here $\tilde{\W}_{i\pm 1/2}^p$ is a 
TVD-limited version of $\W_{i\pm 1/2}^p$. 
In this work, we consider the min-mod limiters, we refer to \cite{leveque1997wave} for more details. 
If no limiters are used 
($\tilde \W_{i\pm 1/2}^p = \W_{i\pm 1/2}^p$), \eqref{second-order_via_fluct} becomes
the standard second-order Lax-Wendroff scheme, which in terms of fluctuation-like quantities 
is given by
\begin{align*}
  u_i^{n+1}=u_i^n-\frac{\Delta t}{\Delta x}
\left[
  \frac{1}{2}\left(s_{i-1/2}^p+\frac{\Delta t}{\Delta x}|s_{i-1/2}^p|^2\right)\W_{i-1/2}^p
  +
  \frac{1}{2}\left(s_{i+1/2}^p-\frac{\Delta t}{\Delta x}|s_{i+1/2}^p|^2\right)\W_{i+1/2}^p
\right].
\end{align*}
To implement method \eqref{second-order_via_fluct}, 
we need to choose a Riemann solver. Although the scheme is second-order 
regardless of the Riemann solver, the accuracy and robustness depends upon the 
Riemann solver. The approximate Riemann solver based on Rusanov's flux is robust and known to be carbuncle free. 
However, it adds large amounts of numerical diffusion that, for relatively coarse grids, 
might dissipate other physical and important features of the solution. If the approximate Riemann solver 
is based on the Roe's average, the amount of numerical diffusion is highly reduced; however, 
such an approach is prone to develop the carbuncle instability for certain problems. 
Many carbuncle fixes to the Roe's solver have been proposed for the Euler equations. 
In the context of the shallow water equations, we review in \S\ref{sec:riemann_solvers} 
the Riemann solver proposed in \cite{kemm2014note}.

@@

For the shallow water equations in two-dimensions, the structure of the exact (and some 
approximated) Riemann solvers contains three waves. This is the case, for example, of the well-known 
Roe's Riemann solver, see for instance \cite{roe1981approximate} and 
\cite[\S 15.3.3 and \S 21.7]{leveque2002finite}. 
Simpler approaches are also possible. 
For instance, methods \eqref{first-order_via_fluct} and \eqref{second-order_via_fluct}
with the Rusanov's numerical flux are based on only two waves that 
propagate with the same speed in opposite directions.
We refer to \cite{ketcheson2020riemann} and references therein for 
a full and detailed review of these two Riemann solvers.

\subsection{Roe's average}\label{sec:roe}
The Roe's Riemann solver is based on evaluating the flux Jacobian using the Roe's average
\begin{align}\label{roe_average}
  \bar h_{i-1/2}=\frac{1}{2}(h_{i-1}+h_i), \qquad 
  \hat u_{i-1/2}=\frac{\sqrt{h_{i-1}}u_{i-1}+\sqrt{h_i}u_i}{\sqrt{h_{i-1}}+\sqrt{h_i}}, \qquad
  \hat v_{i-1/2}=\frac{\sqrt{h_{i-1}}v_{i-1}+\sqrt{h_i}v_i}{\sqrt{h_{i-1}}+\sqrt{h_i}}.
\end{align}
Let 
\begin{align*}
\hat A_{i-1/2} =
\begin{bmatrix}
  0 & 1 & 0 \\
  -\hat u_{i-1/2}^2+g\bar h_{i-1/2} & 2\hat u_{i-1/2} & 0 \\
  -\hat u_{i-1/2} \hat v_{i-1/2} & \hat v_{i-1/2} & \hat u_{i-1/2}
\end{bmatrix}
\end{align*}
be the flux Jacobian evaluated using the Roe's average at the interface $i-1/2$.
The average \eqref{roe_average} is designed to guarantee: 
i)  consistency of the approximate flux Jacobian 
($\hat A_{i-1/2}\rightarrow \bff^\prime(\bfu(x_{i-1/2}))$ as $\bfu_i, \bfu_{i-1}\rightarrow \bfu(x_{i-1/2})$); 
ii) hyperbolicity ($\hat A_{i-1/2}$ is diagonalizable with real eigenvalues);
and iii) conservation ($\hat A_{i-1/2}(\bfu_i-\bfu_{i-1})=\bff(\bfu_i)-\bff(\bfu_{i-1})$). 
Let
\begin{align*}
  \hat \bfr^1_{i-1/2} = 
  \begin{bmatrix}
    1 \\ 
    \hat u_{i-1/2}-\sqrt{g\bar h_{i-1/2}}\\
    \hat v_{i-1/2}
  \end{bmatrix},
  \qquad 
  \hat \bfr^2_{i-1/2} = 
  \begin{bmatrix}
    0\\ 
    0\\ 
    1
  \end{bmatrix},
  \qquad
  \hat \bfr^3_{i-1/2} = 
  \begin{bmatrix}
    1 \\ 
    \hat u_{i-1/2}+\sqrt{g\bar h_{i-1/2}}\\
    \hat v_{i-1/2}
  \end{bmatrix}
\end{align*}
be the eigenvectors of $\hat A_{i-1/2}$.
The waves for the two-dimensional Riemann solver based on the Roe's average are 
$\W_{i-1/2}^p=\alpha_{i-1/2}^p\hat r_{i-1/2}^p$, with $p=1,2,3$. Here $\alpha_{i-1/2}^p$ is given by
\begin{align}\label{system_for_alphas}
  \left[\hat \bfr^1_{i-1/2} ~\hat \bfr^2_{i-1/2} ~\hat \bfr^3_{i-1/2}\right]
  \begin{bmatrix}
    \alpha^1_{i-1/2} \\
    \alpha^2_{i-1/2} \\
    \alpha^3_{i-1/2}
  \end{bmatrix}
  =\Delta Q_{i-1/2}:=\bfu_i-\bfu_{i-1}.
\end{align} 
%where $\alpha_{i-1/2}^p$ is the $p$-th entry from the solution of 
%\end{align*}
The speed of the waves is given by the eigenvalues of $\hat A_{i-1/2}$
\begin{align*}
  \hat\lambda_{i-1/2}^1=\hat u_{i-1/2}-g\sqrt{g\bar h_{i-1/2}}, \qquad
  \hat\lambda_{i-1/2}^2=\hat u_{i-1/2}, \qquad
  \hat\lambda_{i-1/2}^3=\hat u_{i-1/2}+g\sqrt{g\bar h_{i-1/2}}. \\
\end{align*}

@@

We close this section by remarking that the fluctuations associated with Rusanov's solver 
can be expressed in terms of waves and speeds from the Roe's solver. To see this, first consider \eqref{rus_cons}
and rewrite the right-going fluctuation as follows:
\begin{align}\label{rus_rs_aux}
  \A^{+,\Rus}\Delta Q_{i-1/2}:=\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\max}
  = \bff(\bfu_{i})-\bff(\bfu_{i-1}) + \lambda_{i-1/2}^{\max}\left(\W_{i-1/2}^{1,\Rus}+\W_{i-1/2}^{2,\Rus}\right)
  -\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\Rus}.
  %\A^{+,\Rus}\Delta Q_{i-1/2}=\frac{1}{2}\sum_p\left(\hat \lambda_{i-1/2}^p + \lambda_{i-1/2}^{\max}\right)\W_{i-1/2}^p
\end{align}
From \eqref{system_for_alphas} and \eqref{rusanov_waves} we get 
$\W^{1,\Rus}_{i-1/2}+\W^{2,\Rus}_{i-1/2}=\bfu_i-\bfu_{i-1}=\sum_p\W_{i-1/2}^p$. 
Using \eqref{cons_roe}, \eqref{rus_rs_aux} becomes 
\begin{align*}
  \lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\max}
  = \sum_p \left(\hat \lambda_{i-1/2}^p + \lambda_{i-1/2}^{\max}\right)\W_{i-1/2}^p
  -\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\Rus},
\end{align*}
which implies that 
\begin{subequations}\label{fluct_rus}
\begin{align}
  \A^{+,\Rus}\Delta Q_{i-1/2}=\frac{1}{2}\sum_p\left(\hat \lambda_{i-1/2}^p + \lambda_{i-1/2}^{\max}\right)\W_{i-1/2}^p,
\end{align}
and similarly, 
\begin{align}
  \A^{-,\Rus}\Delta Q_{i+1/2}=\frac{1}{2}\sum_p\left(\hat \lambda_{i+1/2}^p - \lambda_{i+1/2}^{\max}\right)\W_{i+1/2}^p.
\end{align}
\end{subequations}

@@

\section{Entropy-dissipative Riemann solver}\label{sec:ent_diss_rs}
Consider a hyperbolic system of conservation laws 
\begin{align}\label{cons_law}
  \frac{\partial \bfu(\bfx)}{\partial t} + \nabla\cdot \bff(\bfu(\bfx))=0, \quad \bfx\in\Omega,
\end{align}
with a en entropy pair $\left(\eta(\bfu), \bfq(\bfu)\right)$. 
A weak solution of \eqref{cons_law} is called an entropy solution if 
\begin{align}\label{ent_inequality}
  \frac{\partial \eta(\bfu(\bfx))}{\partial t} + \nabla\cdot \bfq(\bfu(\bfx))\leq 0, \quad \bfx\in\Omega,
\end{align}
holds, in a distribution sense, for any entropy pair.
It is desirable for a numerical method to satisfy a discrete version of the entropy inequality 
\eqref{ent_inequality}, but many widely-used discretizations do not.
In general, an entropy stable scheme introduces nonlinear stabilization, which is 
particularly important for solutions with strong discontinuities that might converge to non-physical 
weak solutions. 
In this work, we propose an entropy stable approximate Riemann solver for the one dimensional 
shallow water equations. We use this Riemann solver as building block within the Lax-Wendroff-LeVeque 
method, as described in \cite{leveque1997wave,leveque2002finite} and implemented in Clawpack.

\subsection{Entropy residual}
Let $S_i$ be the set of cells containing cell $i$ and those that share a face with it. 
Also let $\bfv(\bfu)=\eta^\prime(\bfu)$ be the entropy variable. 
Based on \cite{guermond2011entropy}, we consider 
\begin{align}\label{ent_residual}
  \int_{S_i} \bfv(\bfu) \cdot \left[ \frac{\partial \bfu}{\partial t} + \nabla\cdot \bff(\bfu)\right]d\bfx
  =\int_{S_i} \left[\frac{\partial \eta(\bfu)}{\partial t} + \bfv(\bfu) \cdot \nabla\cdot \bff(\bfu)\right] d\bfx
\end{align}
as a measurement of the entropy production in $S_i$. 
To avoid approximating the time derivative of the entropy, we follow 
\cite{guermond2018second, guermond2018well} and use 
$\int_{K_i} \frac{\partial\eta(\bfu)}{\partial t}d\bfx=-\int_{K_i} \nabla\cdot\bfq(\bfu) d\bfx$, 
which holds in smooth regions. Then we define 
\begin{align}\label{Ri}
  R_i := \frac{N_i}{D_i}, 
\end{align}
with
\begin{align*}
  N_i=
  \left|\int_{S_i} \left[\bfv(\bfu) \cdot \nabla\cdot \bff(\bfu) - \nabla\cdot\bfq(\bfu) \right] d\bfx\right|,
\end{align*}
and $D_i$ being an upper bound of $N_i$ that acts as normalization so that $0\leq R_i\leq 1$.
Note that $N_i\approx 0$ if $\bfu$ is smooth in $S_i$. 
In our implementation, we use
\begin{align*}
  N_i\approx \left|\bfv(\bfu_i)\cdot \int_{S_i}\nabla\cdot \bff(\bfu)d\bfx 
  -\int_{S_i}\nabla\cdot\bfq(\bfu) d\bfx\right|,
  \quad 
  %D_i = ||\bfv(\bfu_i)||_{\ell^2}\left|\left|\int_{S_i}\nabla\cdot\bff(\bfu)d\bfx\right|\right|_{\ell^2}
  %+\left|\int_{S_i}\nabla\cdot\bfq(\bfu)d\bfx\right|.
  D_i = \sum_{k=1}^{d+1}|\bfv_k(\bfu_i)|\left|\int_{S_i}\left(\nabla\cdot\bff(\bfu)\right)_kd\bfx\right|
  +\left|\int_{S_i}\nabla\cdot\bfq(\bfu)d\bfx\right|,
\end{align*}
where $({\bf z})_k$ denotes the $k$-th component of ${\bf z}\in\mathbb{R}^{d+1}$ and $d$ is the number of physical 
dimensions. In the next section, we use $R_i$ to blend Rusanov's and Roe's Riemann solvers.

>>>>

\section{Numerical methods}
In this work, we consider Riemann-based finite volume methods.
Considering a one dimensional hyperbolic conservation law,
the cell average of the numerical solution given by a
finite volume scheme (with forward Euler) can be written as follows:
\begin{align}\label{first-order_FV}
  \bfu_i^{n+1}=\bfu_i^n-\frac{\Delta t}{\Delta x}\left[\bfF_{i+1/2}-\bfF_{i-1/2}\right],
\end{align}
where $\bfF_{i+1/2}=\bfF_{i+1/2}(u^n)$ is a numerical flux; i.e., $\bfF_{i+1/2}$ is
a discretization of the flux $\bff(\bfu(x_{i+1/2},t^n))$.
This numerical flux is composed of two parts, a consistent term that approximates
the physical flux at the interface between cells $i$ and $i+1$, and a stabilization term.
For instance, the well-known Rusanov's numerical flux is given by
\begin{align}\label{rusanov_num_flux}
  \bfF_{i+1/2}(\bfu)=\frac{\bff(\bfu_{i+1})+\bff(\bfu_i)}{2} - \lambda_{i+1/2}(\bfu_{i+1}-\bfu_i),
\end{align}
where $\lambda_{i+1/2}$ is an upper bound on the wave speed associated with the Riemann problem
composed of the states $\bfu_i$ and $\bfu_{i+1}$.
%If $z=u$ in \eqref{rusanov_num_flux}
%; i.e., if the Rusanov's numerical flux is obtained considering
%the cell averages of the solution
%and if the time step is small enough,
Method \eqref{first-order_FV} with Rusanov's flux \eqref{rusanov_num_flux}
is positivity preserving and entropy stable; see for instance
\cite{perthame1996positivity,tadmor2003entropy,guermond2016invariant}
However, the accuracy of the numerical flux, and hence of the method, is
limited to first order.
Two alternatives are common to improve the accuracy properties of Rusanov's method.
The numerical flux $\bfF_{i+1/2}$ can be approximated based on high-order polynomial
reconstructions (or combinations of them) around the interface $i+1/2$.
For smooth solutions, this strategy leads to arbitrarily high-order
methods. Alternatively, second-order accuracy can be obtained by removing some of the
numerical dissipation introduced by the stabilization component of the numerical flux.
The second-order Lax-Wendroff method \cite{lax1959systems} is a popular example of this approach.
In this work we consider an adaptation of Lax-Wendroff method,
sometimes known as Lax-Wendroff-LeVeque \cite{leveque1997wave,leveque2002finite},
that incorporates limiters into the second-order correction.
These high-order corrections to Rusanov's method require careful treatment to guarantee
the solution is still positivity preserving and/or entropy stable.

The accuracy of a given first- or second-order method depends heavily on the Riemann solver.
In particular, Rusanov's Riemann solver is highly dissipative. In contrast, the well-known
Roe's Riemann solver produces more accurate solutions.
%
As we discussed in the introduction, finite volume methods with
Roe's Riemann solvers are prone to the carbuncle instability for the CHJ.
In contrast, if we use Rusanov's
Riemann solver, no carbuncles appear; however, many other important and physical features of
the solution are dissipated, unless the grid is highly refined.
Our main objective is to study the CHJ and to propose a Riemann solver
that is carbuncle free but do not dissipate important features of the solution.
To do that we propose a Riemann solver that blends Rusanov's and Roe's solvers.
Our blending criterion is based on the residual of the entropy (in)equality,
which induces dissipation of entropy near discontinuities.
Blending these Riemann solvers has been proposed before
to solve the Euler equations, see e.g.
\cite{nishikawa2008very,wang2016developing,jaisankar2007diffusion,ohwada2018simple,deng2019new,ray2013entropy},
the shallow water equations, see e.g. \cite{bader2014carbuncle,kemm2014note},
and even the Navier-Stokes equations, see e.g. \cite{nishikawa2008very,ohwada2018simple}.
In these references, the authors use blending criteria different than ours.
Nevertheless, fixing the carbuncle instability by imposing entropy stability is not new.
For example, in \cite{ismail2009affordable,ismail2009proposed},
the authors developed carbuncle free solvers for the Euler equations
via dissipation of entropy.

@@

The second-order Lax-Wendroff-LeVeque method removes some of the numerical dissipation in
\eqref{first-order_via_fluct} via a flux correction at each interface. The method is written as follows:
\begin{align}\label{second-order_via_fluct}
  u_i^{n+1} = u_i^n-\frac{\Delta t}{\Delta x}\left[\A^+\Delta Q_{i-1/2}+\A^-\Delta Q_{i+1/2}\right]
  -\frac{\Delta t}{\Delta x}\left[\tilde{F}_{i+1/2}-\tilde{F}_{i-1/2}\right],
\end{align}
where
\begin{align*}
  \tilde{F}_{i\pm 1/2}=\frac{1}{2}\sum_p|s_{i\pm 1/2}^p|\left(1-\frac{\Delta t}{\Delta x}|s_{i\pm 1/2}^p|\right)\tilde\W_{i\pm 1/2}^p
\end{align*}
are the flux corrections at the interfaces $i\pm 1/2$. Here $\tilde{\W}_{i\pm 1/2}^p$ is a
TVD-limited version of $\W_{i\pm 1/2}^p$.
In this work, we consider the min-mod limiters, we refer to \cite{leveque1997wave} for more details.
If no limiters are used
($\tilde \W_{i\pm 1/2}^p = \W_{i\pm 1/2}^p$), \eqref{second-order_via_fluct} becomes
the standard second-order Lax-Wendroff scheme, which in terms of fluctuation-like quantities
is given by
\begin{align*}
  u_i^{n+1}=u_i^n-\frac{\Delta t}{\Delta x}
\left[
  \frac{1}{2}\left(s_{i-1/2}^p+\frac{\Delta t}{\Delta x}|s_{i-1/2}^p|^2\right)\W_{i-1/2}^p
  +
  \frac{1}{2}\left(s_{i+1/2}^p-\frac{\Delta t}{\Delta x}|s_{i+1/2}^p|^2\right)\W_{i+1/2}^p
\right].
\end{align*}
To implement method \eqref{second-order_via_fluct},
we need to choose a Riemann solver. Although the scheme is second-order
regardless of the Riemann solver, the accuracy and robustness depends upon the
Riemann solver. The approximate Riemann solver based on Rusanov's flux is robust and known to be carbuncle free.
However, it adds large amounts of numerical diffusion that, for relatively coarse grids,
might dissipate other physical and important features of the solution. If the approximate Riemann solver
is based on Roe's average, the amount of numerical diffusion is highly reduced; however,
such an approach is prone to develop the carbuncle instability for certain problems.
Many carbuncle fixes to Roe's solver have been proposed for the Euler equations.
In the context of the shallow water equations, we review in \S\ref{sec:riemann_solvers}
the Riemann solver proposed in \cite{kemm2014note}.

@@

For the shallow water equations in two-dimensions, the structure of the exact (and some
approximated) Riemann solvers contains three waves. This is the case, for example, of the well-known
Roe's Riemann solver, see for instance \cite{roe1981approximate} and
\cite[\S 15.3.3 and \S 21.7]{leveque2002finite}.
Simpler approaches are also possible.
For instance, methods \eqref{first-order_via_fluct} and \eqref{second-order_via_fluct}
with Rusanov's numerical flux are based on only two waves that
propagate with the same speed in opposite directions.
We refer to \cite{ketcheson2020riemann} and references therein for
a full and detailed review of these two Riemann solvers.

\subsection{Roe's average}\label{sec:roe}
The Roe's Riemann solver is based on evaluating the flux Jacobian using Roe's average
\begin{align}\label{roe_average}
  \bar h_{i-1/2}=\frac{1}{2}(h_{i-1}+h_i), \qquad
  \hat u_{i-1/2}=\frac{\sqrt{h_{i-1}}u_{i-1}+\sqrt{h_i}u_i}{\sqrt{h_{i-1}}+\sqrt{h_i}}, \qquad
  \hat v_{i-1/2}=\frac{\sqrt{h_{i-1}}v_{i-1}+\sqrt{h_i}v_i}{\sqrt{h_{i-1}}+\sqrt{h_i}}.
\end{align}
Let
\begin{align*}
\hat A_{i-1/2} =
\begin{bmatrix}
  0 & 1 & 0 \\
  -\hat u_{i-1/2}^2+g\bar h_{i-1/2} & 2\hat u_{i-1/2} & 0 \\
  -\hat u_{i-1/2} \hat v_{i-1/2} & \hat v_{i-1/2} & \hat u_{i-1/2}
\end{bmatrix}
\end{align*}
be the flux Jacobian evaluated using Roe's average at the interface $i-1/2$.
The average \eqref{roe_average} is designed to guarantee:
i)  consistency of the approximate flux Jacobian
($\hat A_{i-1/2}\rightarrow \bff^\prime(\bfu(x_{i-1/2}))$ as $\bfu_i, \bfu_{i-1}\rightarrow \bfu(x_{i-1/2})$);
ii) hyperbolicity ($\hat A_{i-1/2}$ is diagonalizable with real eigenvalues);
and iii) conservation ($\hat A_{i-1/2}(\bfu_i-\bfu_{i-1})=\bff(\bfu_i)-\bff(\bfu_{i-1})$).
Let
\begin{align*}
  \hat \bfr^1_{i-1/2} =
  \begin{bmatrix}
    1 \\
    \hat u_{i-1/2}-\sqrt{g\bar h_{i-1/2}}\\
    \hat v_{i-1/2}
  \end{bmatrix},
  \qquad
  \hat \bfr^2_{i-1/2} =
  \begin{bmatrix}
    0\\
    0\\
    1
  \end{bmatrix},
  \qquad
  \hat \bfr^3_{i-1/2} =
  \begin{bmatrix}
    1 \\
    \hat u_{i-1/2}+\sqrt{g\bar h_{i-1/2}}\\
    \hat v_{i-1/2}
  \end{bmatrix}
\end{align*}
be the eigenvectors of $\hat A_{i-1/2}$.
The waves for the two-dimensional Riemann solver based on Roe's average are
$\W_{i-1/2}^p=\alpha_{i-1/2}^p\hat r_{i-1/2}^p$, with $p=1,2,3$. Here $\alpha_{i-1/2}^p$ is given by
\begin{align}\label{system_for_alphas}
  \left[\hat \bfr^1_{i-1/2} ~\hat \bfr^2_{i-1/2} ~\hat \bfr^3_{i-1/2}\right]
  \begin{bmatrix}
    \alpha^1_{i-1/2} \\
    \alpha^2_{i-1/2} \\
    \alpha^3_{i-1/2}
  \end{bmatrix}
  =\Delta Q_{i-1/2}:=\bfu_i-\bfu_{i-1}.
\end{align}
%where $\alpha_{i-1/2}^p$ is the $p$-th entry from the solution of
%\end{align*}
The speed of the waves is given by the eigenvalues of $\hat A_{i-1/2}$
\begin{align*}
  \hat\lambda_{i-1/2}^1=\hat u_{i-1/2}-g\sqrt{g\bar h_{i-1/2}}, \qquad
  \hat\lambda_{i-1/2}^2=\hat u_{i-1/2}, \qquad
  \hat\lambda_{i-1/2}^3=\hat u_{i-1/2}+g\sqrt{g\bar h_{i-1/2}}. \\
\end{align*}

@@

We close this section by remarking that the fluctuations associated with Rusanov's solver
can be expressed in terms of waves and speeds from Roe's solver. To see this, first consider \eqref{rus_cons}
and rewrite the right-going fluctuation as follows:
\begin{align}\label{rus_rs_aux}
  \A^{+,\Rus}\Delta Q_{i-1/2}:=\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\max}
  = \bff(\bfu_{i})-\bff(\bfu_{i-1}) + \lambda_{i-1/2}^{\max}\left(\W_{i-1/2}^{1,\Rus}+\W_{i-1/2}^{2,\Rus}\right)
  -\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\Rus}.
  %\A^{+,\Rus}\Delta Q_{i-1/2}=\frac{1}{2}\sum_p\left(\hat \lambda_{i-1/2}^p + \lambda_{i-1/2}^{\max}\right)\W_{i-1/2}^p
\end{align}
From \eqref{system_for_alphas} and \eqref{rusanov_waves} we get
$\W^{1,\Rus}_{i-1/2}+\W^{2,\Rus}_{i-1/2}=\bfu_i-\bfu_{i-1}=\sum_p\W_{i-1/2}^p$.
Using \eqref{cons_roe}, \eqref{rus_rs_aux} becomes
\begin{align*}
  \lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\max}
  = \sum_p \left(\hat \lambda_{i-1/2}^p + \lambda_{i-1/2}^{\max}\right)\W_{i-1/2}^p
  -\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\Rus},
\end{align*}
which implies that
\begin{subequations}\label{fluct_rus}
\begin{align}
  \A^{+,\Rus}\Delta Q_{i-1/2}=\frac{1}{2}\sum_p\left(\hat \lambda_{i-1/2}^p + \lambda_{i-1/2}^{\max}\right)\W_{i-1/2}^p,
\end{align}
and similarly,
\begin{align}
  \A^{-,\Rus}\Delta Q_{i+1/2}=\frac{1}{2}\sum_p\left(\hat \lambda_{i+1/2}^p - \lambda_{i+1/2}^{\max}\right)\W_{i+1/2}^p.
\end{align}
\end{subequations}

@@

\section{Entropy-dissipative Riemann solver}\label{sec:ent_diss_rs}
Consider a hyperbolic system of conservation laws
\begin{align}\label{cons_law}
  \frac{\partial \bfu(\bfx)}{\partial t} + \nabla\cdot \bff(\bfu(\bfx))=0, \quad \bfx\in\Omega,
\end{align}
with a en entropy pair $\left(\eta(\bfu), \bfq(\bfu)\right)$.
A weak solution of \eqref{cons_law} is called an entropy solution if
\begin{align}\label{ent_inequality}
  \frac{\partial \eta(\bfu(\bfx))}{\partial t} + \nabla\cdot \bfq(\bfu(\bfx))\leq 0, \quad \bfx\in\Omega,
\end{align}
holds, in a distribution sense, for any entropy pair.
It is desirable for a numerical method to satisfy a discrete version of the entropy inequality
\eqref{ent_inequality}, but many widely-used discretizations do not.
In this work, we propose an entropy stable approximate Riemann solver for the one dimensional
shallow water equations. We use this Riemann solver as building block within the Lax-Wendroff-LeVeque
method, as described in \cite{leveque1997wave,leveque2002finite} and implemented in Clawpack.

\subsection{Entropy residual}
Let $S_i$ be the set of cells containing cell $i$ and those that share a face with it.
Also let $\bfv(\bfu)=\eta^\prime(\bfu)$ be the entropy variable.
Based on \cite{guermond2011entropy}, we consider
\begin{align}\label{ent_residual}
  \int_{S_i} \bfv(\bfu) \cdot \left[ \frac{\partial \bfu}{\partial t} + \nabla\cdot \bff(\bfu)\right]d\bfx
  =\int_{S_i} \left[\frac{\partial \eta(\bfu)}{\partial t} + \bfv(\bfu) \cdot \nabla\cdot \bff(\bfu)\right] d\bfx
\end{align}
as a measurement of the entropy production in $S_i$.
To avoid approximating the time derivative of the entropy, we follow
\cite{guermond2018second, guermond2018well} and use
$\int \frac{\partial\eta(\bfu)}{\partial t}d\bfx=-\int \nabla\cdot\bfq(\bfu) d\bfx$,
which holds in smooth regions. Then we define
\begin{align}\label{Ri}
  \theta_i := \frac{R_i}{D_i},
\end{align}
with
\begin{align*}
  R_i=
  \left|\int_{S_i} \left[\bfv(\bfu) \cdot \nabla\cdot \bff(\bfu) - \nabla\cdot\bfq(\bfu) \right] d\bfx\right|,
\end{align*}
and $D_i$ being an upper bound of $R_i$ that acts as normalization so that $0\leq \theta_i\leq 1$.
Note that $R_i\approx 0$ if $\bfu$ is smooth in $S_i$.
In our implementation, we use
\begin{align*}
  R_i\approx \left|\bfv(\bfu_i)\cdot \int_{S_i}\nabla\cdot \bff(\bfu)d\bfx
  -\int_{S_i}\nabla\cdot\bfq(\bfu) d\bfx\right|,
  \quad
  %D_i = ||\bfv(\bfu_i)||_{\ell^2}\left|\left|\int_{S_i}\nabla\cdot\bff(\bfu)d\bfx\right|\right|_{\ell^2}
  %+\left|\int_{S_i}\nabla\cdot\bfq(\bfu)d\bfx\right|.
  D_i = \sum_{k=1}^{d+1}|\bfv_k(\bfu_i)|\left|\int_{S_i}\left(\nabla\cdot\bff(\bfu)\right)_kd\bfx\right|
  +\left|\int_{S_i}\nabla\cdot\bfq(\bfu)d\bfx\right|,
\end{align*}
where $({\bf z})_k$ denotes the $k$-th component of ${\bf z}\in\mathbb{R}^{d+1}$ and $d$ is the number of physical
dimensions. In the next section, we use $\theta_i$ to blend Rusanov's and Roe's Riemann solvers.

\subsection{Blended Riemann solver}\label{sec:blended_rs}
In this section, we use \eqref{Ri}, which is a normalized measurement of the entropy residual \eqref{ent_residual},
to blend Rusanov's and Roe's Riemann solvers. To do this, we define the following numerical flux:
\begin{align}\label{num_flux_ev}
  F_{i+1/2} = \frac{\bff(\bfu_{i+1})+\bff(\bfu_i)}{2}
  - \frac{1}{2} \left( \theta_{i+1/2}\lambda_{i+1/2}^{\max} I + (1-\theta_{i+1/2})|\hat A_{i+1/2}| \right)(\bfu_{i+1}-\bfu_{i}),
\end{align}
and similarly for $F_{i-1/2}$. Here $I\in\mathbb{R}^{3\times 3}$ is the identity matrix,
$\theta_{i+1/2}=\max\{\theta_{i+1},\theta_i\}$, $\lambda_{i+1/2}^{\max}$ is the
upper bound on the wave speed used in Rusanov's Riemann solver (see \S \ref{sec:rusanov}),
and $\hat A_{i+1/2}$ is Roe's averaged flux Jacobian (see \S \ref{sec:roe}).
